A Linear Technique to Understand Non-Normal Turbulence Applied to a Magnetized Plasma

### 2014-02-24 Referee_Response1/referee_response.tex

\begin{document}

{\bf Response to Referees \\
 \\
 \\} 
Brett Friedman \\
Updated: \today \\

\hrulefill

\section{Overview of Changes}

Before responding to specific comments, questions, and requests from the referees, we would like to note that we have significantly changed parts of the manuscript. First, we have
changed the way we calculate the $|e^{A t}u(0)|$ curves where we used an ensemble of initial conditions. Before we made many initial conditions, evolved them by this formula and then
calculated the ensemble average. This way seemed a bit like performing numerical simulations. Now we have an analytic method of calculation which doesn't involve setting up any initial
conditions, so it is not at all like simulation (see Eq. 10) and is much more elegant.

Second, recall that our method involved choosing a linear time scale as a proxy for the eddy turnover time, for which we chose the inverse of a linear eigenvalue frequency. 
As the first referee pointed out, we didn't properly justify this choice. Since we have no theoretical justification, 
we have now included a few more linear time scales and studied the result for each of these. Furthermore, we no longer claim that our method
is predictive because we have no way to know \emph{a priori} which linear time scale will work best. However, all of the linear time scales do work well to a certain degree, so
we do obtain a degree of predictivity.

Otherwise, we have made various minor changes and changed two of the figures to make sense with the changes mentioned above. 
Overall the manuscript is significantly different than before, and we feel it is much stronger.

@@

\section{Referee 1}

\emph{1) After advocating the use of non-modal structures rather than linear eigenmodes for the
entirety of the paper, the author use a time-scale derived from linear eigenmodes (see
page 4, paragraph 2). Furthermore, they use the frequency for the n=1 mode instead of
that for the n=0 mode(s), with no apparent justification (other than "nonlinear coupling”).
It would seem much more consistent to use the time-scale for the nonlinear instability
they talk about earlier (although this would mean perhaps a loss of predictability that they
are very keen on). To us, this appears to be the weakest link in the paper. Perhaps, if they
present this paper as a study of the turbulence properties instead, they could give
information on how the curves in Figure 3 change with the choice of the linear time-
scale. It would seem that this would be much more useful to the reader in assessing the
power of the non-modal analysis, rather than presenting just two choices, and claiming
predictability. } \\

Response: This is a valid criticism. We thought a lot about this and tried a few different linear time scales. The one we chose worked best, but it is difficult to justify it otherwise.
We agree that this is the weakest link of the paper. We also think that the suggestion of presenting several choices of linear time scales is a good one, so this is what we've done.
We've presented four different choices of linear time scale and compared the results to each other. We have also softened our language on predictive capability and presented our results
as a way of understanding the turbulent growth rate through linear non-modal calculations with the ultimate goal of achieving prediction. \\

@@

\emph{2) What closure assumption is used to write equation (5)?} \\

Response: We do not really understand this question. We do not use any closure to write equation 5. It is an exact equation that can be derived from the model equations~\cite{friedman2012b,friedman2013}. 
We have also now written it differently to make it simpler and more obvious. \\

\emph{3) On page 2, last paragraph, there is a puzzling statement: "Perhaps one of the most
useful statistical properties that we can predict is the turbulent growth rate spectrum.” Do
they actually have a measurement of the turbulent growth rate spectrum? Since
predictability is one of their prime concerns, what measurable transport property does the
growth rate spectrum relate to?} \\

Response: We acknowledge that this was too cryptic a statement. In it's place, we have inserted the following, ``These properties can be calculated through mixing length formulas if the turbulent growth rate spectrum is known.'' Then, as before, we proceed to calculate the turbulent growth rate spectrum from the nonlinear simulation (shown in Figure 1a) and also calculate it from our non-modal linear procedure. \\

@@

\emph{4) We think a log scale should be used for Figure 2. On the scale used in the paper, it is
impossible to determine if the growth rate at short times is somewhat similar to $G_{max} (t)$
or quite different. If similar, the potential for formulating fully analytic models is more
promising.} \\

Response: We have changed Figure 2 so that we do not show $G_{max}(t)$ any longer. The growth rate of the ensemble of curves is not close to the growth rate of $G_{max}(t)$. $G_{max}(t)$  grows so quickly,
and it isn't clear how one would use it in a procedure like ours because the points along it require linear evolution from very specific initial conditions, and every point comes from a different initial condition. \\

@@

\emph{5) Minor point: Figure 3 is confusing simply because the important curves are not
sufficiently well-delineated. For example, the turbulent growth rate spectrum should be
bold (or something equivalent) to distinguish it from the others.} \\

Response: This figure is also much different now, but we have tried to make all of the curves distinguishable by using different colors along with symbols so that they are even distinguishable in black and white.\\

@@

\beq
\diff{\mathbf{u}}{t} = \mathbf{A u} \nonumber
\eeq

\emph{I would expect the prediction of the short time behavior of some norm of $\mathbf{u}$ through the mathematical analysis of $\mathbf{A}$
by methods other than the standard eigenmode decomposition, e.g., by methods based on the $\epsilon$-pseudospectrum and/or numerical range.}

\emph{If I understand correctly, the method presented here starts by dropping the nonlinear terms in
the equations of motion of the reduced Braginskii 2-fluid model. The resulting linear system is
numerically simulated up to a time $\tau_{nl}$ at which the system is then ‘re-randomized’ — whatever
it means. A growth rate $\gamma_{TR}$ defined and computed using data between $t = 0$ and $t = \tau_{nl}$ then
constitutes the prediction by this method. So it seems that the ‘non-modal method’ described in
this manuscript is basically a numerical simulation of the linearized equations of motion. While
prediction of turbulent properties by numerical simulations can be useful, it probably has less
impact than a theoretical prediction (which is admittedly much more difficult).} \\

Response: We contend that our method is a form of 'non-modal analysis'. That is, we use only linear calculations to find growth rates without using normal modes (eigenvalues and eigenvectors). While most non-modal analysis in the literature seems to center around the calculation of the $epsilon$-pseudospectrum, much also includes the calculation of $||e^{\mathbf{A} t}||$ and $||e^{\mathbf{A} t} \mathbf{u}(0)||$ with various optimal $\mathbf{u}(0)$ vectors, i.e. to get the numerical range. More importantly, however, it would be helpful to have a more theoretical prediction that doesn't require simulation. In that regard, we have come across a method~\cite{camargo1998} that can replicate our numerical simulations with a simple calculation that looks much more like a typical non-modal calculation. See Eqs. 9 and 10. \\

@@

\emph{2) For the procedure described above, there are several things I do not fully understand. First of
all, if $\gamma_{TR}$ is only computed using data for $0 < t < \tau_{nl}$, how does the ‘re-randomization’ at $t = \tau_{nl}$
come into play? I am probably missing something here.} \\

Response: We agree that this was a confusing way to word our procedure. We therefore have replaced our previous wording with the following: ``we start with an ensemble of random initial conditions, evolve them linearly for a time $\tau_{\rm{nl}}$, and then take the time and ensemble average growth rate of these curves. 
This procedure mimicks what would happen if the turbulence evolved linearly for a time $\tau_{\rm{nl}}$, randomized, and then repeated.'' Also, we use the non-modal calculation discussed above, added more equations, and replaced Figure 2 to make our procedure more clear.
\\

@@

Finally, the reason we plot $E(m,n,t)$ with the nonlinearities turned off is to show that the linear terms alone cause (transient) growth of the fluctuations despite the fact that the $n=0$ mode numbers contain no unstable eigenmodes. It illustrates that the turbulent structures inject energy into the fluctuations in the short term. \\

\emph{4) I also feel that the clarity of the present manuscript suffers because of the page limitation of
PRL, a lot of the symbols are not defined (e.g. $v_E$, $N_r$ ) making it difficult for the general audience.
Together with the concerns raised above, I am afraid I cannot recommend the publication of the
present manuscript in the PRL.} \\

Response: We have now tried to define all symbols and have also removed a lot of unnecessary parts of the paper and replaced them with better text. We hope the major changes have made the manuscript significantly better and more valuable. \\

### 2014-02-24 Transient_growth_prl.tex
Feb 23

\begin{abstract}
In nonlinear dynamical systems with highly nonorthogonal linear eigenvectors, linear non-modal analysis is more useful than normal mode analysis in predicting turbulent properties. 
However, the non-trivial time evolution of non-modal structures makes quantitative prediction difficult. We present a technique to overcome this difficulty by modelling the effect that the advective nonlinearities have on spatial turbulent structures. The nonlinearities are taken as a periodic randomizing force with
period consistent with critical balance arguments. We apply this technique to a model of drift wave turbulence in the Large Plasma Device (LAPD) 
[W. Gekelman \emph{et al.}, Rev. Sci. Inst. {\bf 62}, 2875 (1991)], where non-modal effects dominate the turbulence.
We compare the resulting growth rate spectra to that obtained from a nonlinear simulation, showing good qualitative agreement.
\end{abstract}

@@

Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
still heavily relies on normal mode analysis to inform turbulent predictions and observations, with a few notable exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing an approach to understand turbulent properties using only non-modal linear calculations with the goal of 
ultimately making quantitative predictions. 
Our approach is to evolve the linear solution from an ensemble of random initial conditions 
and calculate the average growth rate of the solutions over a specified timescale.  Optimally this time scale would be the
nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic \emph{linear} time. Since several linear time scales exist, we test them all and compare the results to determine which works best.
The procedure ultimately produces an effective growth rate spectrum that can be used to predict turbulent properties such as saturation levels through mixing length arguments.
While the concepts behind this technique are general enough to be applied to many nonlinear dynamical systems, the details vary for each case, 
so we restrict our treatment to one particular turbulence model. For this model, the technique qualitatively reproduces the turbulent growth rate spectrum of the direct 
nonlinear simulation fairly well, especially in comparison to the linear eigenmode spectrum.

@@

The equations are global and we retain only advective nonlinearities, which are written with Poisson brackets. 
We add artificial diffusion and viscosity terms with small numerical coefficients ($10^{-3}$)
to ensure numerical stability in nonlinear simulations, which are performed with the BOUT++ code~\cite{dudson2009}. We use periodic axial and zero value radial 
boundary conditions. Further details of the model, including validation studies, may be found in the references~\cite{Popovich2010a,Popovich2010b,Umansky2011,friedman2012b,friedman2013},
though we mention here that the model does very well in reproducing the statistical properties of the experimentally-observed turbulence.

@@

\beq
\label{dEdt_def}
\diff{E(m,n)}{t} = \diff{E_{l}(m,n)}{t} + \diff{E_{nl}(m,n)}{t}
\eeq
where $m$ and $n$ represent the azimuthal and axial Fourier mode numbers. 
$\diff{E_{l}(m,n)}{t}$ comes from the linear terms in Eqs.~\ref{ni_eq}-\ref{te_eq}, $\diff{E_{nl}(m,n)}{t}$ comes from the nonlinear terms. 
$\diff{E_{l}(m,n)}{t}$ describes the injection (or dissipation) of energy into the fluctuations from the free energy in the equilibrium gradients.
$\diff{E_{nl}(m,n)}{t}$ describes the energy exchange between fluctuations with different $m,n$ and it is conservative: $\sum_{m,n} \diff{E_{nl}(m,n)}{t} = 0$.
Moreover, in quasi-steady state turbulence, the rate of energy injection (or dissipation) into the fluctuations at each $m,n$ by the linear terms must be balanced by
the rate of energy removal (or deposition) from the nonlinear terms:

@@

From, Eqs.~\ref{dEdt_def} and~\ref{steady_state}, it follows that $ \gamma(m,n) = \gamma_l(m,n) + \gamma_{nl}(m,n) = 0$.
The rate of energy injection $\gamma_l(m,n)$ from the equilibrium gradient into the fluctuations is a quantity of great interest~\cite{friedman2012b,terry2006b}. 
When positive for some wavenumber, turbulence can be sustained. Additionally, it may be used in the mixing length formula $\gamma/k_\perp^2$ to predict the turbulent saturation level, 
and it is proportional to the transport rate~\cite{terry2006b}. 
$\gamma_l(m,n)$ may be calculated from the spatial structures of of the plasma state variables, so it has meaning in turbulent plasma states. We plot
$\gamma_l(m,n)$ in Fig.~\ref{n0_n1_growth} a) for both the linear and steady state turbulent stages of our nonlinear simulation. The linear stage, which occurs when fluctuations
are small and exponentially growing has a time-independent $\gamma_l(m,n)$ equal to $\gamma_s(m,n)$ -- the ``spectral'' growth rate of the fastest growing eigenmode at each $m,n$.
During the turbulent stage, $\gamma_l(m,n)$ is time-dependent (indicated by the $1 \sigma$ spread), and generally much different than $\gamma_s(m,n)$. Significantly,
the turbulent $\gamma_l$ is positive at $n=0$ and low $m$ despite the fact that all linear eigenmodes have $\gamma_s < 0$ for $n=0$. This is a manifestation
of non-normality, for in normal systems, $\gamma_l(m,n) \le \gamma_s(m,n)$. Physically, it is the manifestation of the nonlinear
instability, specifically the linear part of the mechanism in which convective filaments drive density filaments from the equilibrium density gradient.

This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise 
streaks~\cite{trefethen1993,krommes1999}.
Both are transient growth processes. We see this in our simulations by following the evolution of the energy of several $m,n$ modes after turning off the nonlinearities in an already
turbulent simulation. A few representative modes are shown in Fig.~\ref{n0_n1_growth} b). The linear transient growth of the filamentary $n=0$ structures is evident as their modes 
grow transiently before decaying exponentially at the rate of their least stable eigenmode. Such behavior is indicative of non-modal behavior. It must be since 
all $n=0$ linear eigenmodes are stable. Notice also that the $n=1,m=20$ mode decays transiently before growing exponentially with growth rate of the most unstable eigenmode. This transient
is another non-modal result.

Since transient growth is a purely linear phenomenon, it has been a goal of researchers to understand and predict the onset of subcritical turbulence using only linear, non-modal calculations. 
In our system, the turbulence is not subcritical in the traditional sense, but it has a subcritical component in the $n=0$ modes. They have no unstable eigenmodes, but still inject energy into the
turbulence -- $\gamma_l(m<50,n=0) > 0$, yet $\gamma_s(m,n=0) < 0$. It is our goal, then, to use linear non-modal calculations to understand this behavior and to move toward predictive capability
of $\gamma_l$.
To accomplish this, we make the anzatz that nonlinearities randomize the turbulent spatial structure at each wavenumber on a time scale of 
one eddy decorrelation time, while the linearities evolve the spatial structures deterministically. From this we can calculate a ``Transient Growth Rate'' spectrum,
which is our prediction of the turbulent growth rate spectrum. To illustrate,
we begin by taking Eqs.~\ref{ni_eq}-\ref{te_eq} and Fourier decomposing in the azimuthal and axial directions.  Then, we discritize in the radial direction and
approximate radial derivatives with finite differences.  The resulting system of equations may be written in matrix form:

@@

In order to use non-modal analysis to calculate growth rates and other measures, one must choose a norm and inner product with which to work. While any choice of
inner product is possible, a physically relevant one such as an energy inner product is generally preferred~\cite{camargo1998,schmid2007,camporeale2010}. 
Recall that the inner product of two vectors may be written $\left< \mathbf{x},\mathbf{y} \right> = \mathbf{y}^{\dagger} \mathbf{M} \mathbf{x}$,
We choose $\mathbf{M}$ so that $||\mathbf{v}||^2 = \left< \mathbf{v},\mathbf{v} \right> = E$. Furthermore, it is convenient in
computations to use the $L_2$-norm, $||\mathbf{u}||_2^2 = \sum_i |u_i|^2$. This can be accomplished through the change of variables $\mathbf{u} = \mathbf{M}^{\frac{1}{2}} \mathbf{v}$.
Then the linear portion of Eq.~\ref{mat_eq} becomes

\beq
\label{lin_eq_A}
\diff{\mathbf{u}}{t} = \mathbf{A} \mathbf{u},  \quad \rm{where} \ \mathbf{A} = \mathbf{M}^{\frac{1}{2}} \mathbf{B}^{-1} \mathbf{C} \mathbf{M}^{-\frac{1}{2}}.
\eeq
The solution of Eq.~\ref{lin_eq_A} is $\mathbf{u}(t) = e^{\mathbf{A} t} \mathbf{u}(0)$, which
depends on the initial condition $\mathbf{u}(0)$ in addition to the spectral properties of $\mathbf{A}$. For purposes of turbulent growth rate prediction, we are interested in
the behavior of $G(t) = E(t)/E(0) = ||\mathbf{u}(t)||^2/||\mathbf{u}(0)||^2$ and its rate of growth.

@@

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{m20n0_E_ev}}
\caption{An ensemble of growth ratio curves (solid gray lines, with the solid black line the ensemble average)
that start with a random initial condition $u(0)$ and evolve under the linear operator.}
\label{m20n0_E_ev}
\end{figure}

It is common practice in normal mode analysis to look for the least stable eigenmode. 
For the non-modal case, it is common to study the properties of $G_{\rm{max}}(t) = ||e^{\mathbf{A} t}||$ because if this is greater at any time, fluctuations may be amplified, 
leading to subcritical turbulence~\cite{trefethen2005,schmid2007}.
However, it can be misleading to study only $G_{\rm{max}}(t)$ when predicting specific properties of turbulence because
$G_{\rm{max}}(t)$ is only the upper envelope of all possible $G(t)$ curves. No one particular initial condition $\mathbf{u}(0)$ evolves along $G_{\rm{max}}(t)$. 
Furthermore, it isn't obvious what kind of spatial structure or structures will come to dominate a turbulent system.
In non-normal systems, unlike in the normal systems, optimal structures don't amplify themselves, rather, they evolve while increasing the total fluctuating energy.

We contend that the key to understanding and prediction turbulent properties through non-modal analysis is to successfully model the effect that the nonlinearities have on the transient linear processes. 
To this effect, we note that the advective nonlinearity in Eq.~\ref{mat_eq} has the form of the state vector divided by a time $\tau_{\rm{nl}} \sim (v_E k_\perp)^{-1}$. This nonlinear
time scale is generally associated with the eddy turnover or decorrelation time. We therefore present a heuristic model of the nonlinearities 
as a randomizing force that acts on this characteristic nonlinear time scale.
Specifically, we start with an ensemble of random initial conditions, evolve them linearly for a time $\tau_{\rm{nl}}$, and then take the time and ensemble average growth rate of these curves.
This procedure mimicks what would happen if the turbulence evolved linearly for a time $\tau_{\rm{nl}}$, randomized, and then repeated. Now, this procedure does not require actual linear
simulations from an ensemble of random initial conditions. Recall that the time evolution of the energy from an initial condition is

@@

\beq
\label{E_t_ensemble_avg}
\left< E(t)/E(0) \right>_{{\rm{ens}}} = \frac{1}{4 N_r} {\rm{trace}} e^{\mathbf{A} t} e^{\mathbf{A}^{\dagger}t}.
\eeq
We show the validity of this statistical averaging by plotting $1000$ ensemble curves in Fig.~\ref{m20n0_E_ev} along with their expected average from Eq.~\ref{E_t_ensemble_avg} and their actual average,
which line up perfectly.
One point to note is that in global equation sets like ours, where we discritize and use finite differences in the radial direction rather than a Fourier decomposition, randomizing $\mathbf{u}(0)$
amounts to setting the initial $k_r$ spectrum to a step function that goes to zero at the Nyquist wavenumber. This means that $\left< E(t)/E(0) \right>_{{\rm{ens}}}$ may depend on $N_r$, so we must be careful in
choosing $N_r$ for this analysis. We choose $N_r$ such that the grid spacing equals $\rho_s$, the general Nyquist wavenumber of drift wave simulations~\cite{scott1992}.

>>>>

\begin{abstract}
In nonlinear dynamical systems with highly nonorthogonal linear eigenvectors, linear non-modal analysis is more useful than normal mode analysis in predicting turbulent properties. 
However, the non-trivial time evolution of non-modal structures makes quantitative understanding and prediction difficult. 
We present a technique to overcome this difficulty by modelling the effect that the advective nonlinearities have on spatial turbulent structures. The nonlinearities are taken as a periodic randomizing force with
time scale consistent with critical balance arguments. We apply this technique to a model of drift wave turbulence in the Large Plasma Device (LAPD) 
[W. Gekelman \emph{et al.}, Rev. Sci. Inst. {\bf 62}, 2875 (1991)], where non-modal effects dominate the turbulence.
We compare the resulting growth rate spectra to that obtained from a nonlinear simulation, showing good qualitative agreement.
\end{abstract}

@@

Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
still heavily relies on normal mode analysis to inform turbulent predictions and observations, with a few notable exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing an approach to understand turbulent properties using only non-modal linear calculations with the goal of 
ultimately making quantitative predictions. 
Our approach is to evolve the linear solution from an ensemble of random initial conditions 
and calculate the average growth rate of the solutions over a specified timescale.  Optimally this time scale would be the
nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic \emph{linear} time. Since several linear time scales exist, we test them all and compare the results to determine which works best.
The procedure ultimately produces growth rate spectra that can be used to predict turbulent properties such as saturation levels and transport rates through mixing length arguments.
While the concepts behind this technique are general enough to be applied to many nonlinear dynamical systems, the details vary for each case, 
so we restrict our treatment to one particular turbulence model. For this model, the technique qualitatively reproduces the turbulent growth rate spectrum of the direct 
nonlinear simulation fairly well, especially in comparison to the linear eigenmode spectrum.

@@

The equations are global and we retain advective nonlinearities, which are written with Poisson brackets, but we neglect other nonlinear terms. 
We add artificial diffusion and viscosity terms with small numerical coefficients ($10^{-3}$)
to ensure numerical stability in nonlinear simulations, which are performed with the BOUT++ code~\cite{dudson2009}. We use periodic axial and zero value radial 
boundary conditions. Further details of the model, including validation studies, may be found in the references~\cite{Popovich2010a,Popovich2010b,Umansky2011,friedman2012b,friedman2013},
though we mention here that the model does very well in reproducing the statistical properties of the experimentally-observed turbulence.

@@

\beq
\label{dEdt_def}
\diff{E(m,n)}{t} = \diff{E_{l}(m,n)}{t} + \diff{E_{nl}(m,n)}{t}
\eeq
where $m$ and $n$ represent the azimuthal and axial Fourier mode numbers. 
$\diff{E_{l}(m,n)}{t}$ comes from the linear terms in Eqs.~\ref{ni_eq}-\ref{te_eq}. $\diff{E_{nl}(m,n)}{t}$ comes from the nonlinear terms. 
$\diff{E_{l}(m,n)}{t}$ represents the injection (or dissipation) of energy into the fluctuations from the free energy in the equilibrium gradients.
$\diff{E_{nl}(m,n)}{t}$ accounts for the energy exchange between fluctuations with different $m,n$ and it is conservative: $\sum_{m,n} \diff{E_{nl}(m,n)}{t} = 0$.
Moreover, in quasi-steady state turbulence, the rate of energy injection (or dissipation) into the fluctuations at each $m,n$ by the linear terms must be balanced by
the rate of energy removal (or deposition) from the nonlinear terms:

@@

From, Eqs.~\ref{dEdt_def} and~\ref{steady_state}, it follows that $ \gamma(m,n) = \gamma_l(m,n) + \gamma_{nl}(m,n) = 0$.
The rate of energy injection $\gamma_l(m,n)$ from the equilibrium gradient into the fluctuations is a quantity of great interest~\cite{friedman2012b,terry2006b}. 
When positive for some wavenumber, turbulence can be sustained. Additionally, it may be used in the mixing length formula $\gamma/k_\perp^2$ to predict the turbulent saturation level, 
and it is proportional to the transport rate~\cite{terry2006b}. 
$\gamma_l(m,n)$ may be calculated from the spatial structures of the plasma state variables, so it is always well-defined, even in a turbulent plasma. We plot
$\gamma_l(m,n)$ in Fig.~\ref{n0_n1_growth} a) for both the linear and steady state turbulent stages of our nonlinear simulation. The linear stage, which occurs when fluctuations
are small and exponentially growing has a time-independent $\gamma_l(m,n)$ equal to $\gamma_s(m,n)$ -- the ``spectral'' growth rate of the fastest growing eigenmode at each $m,n$.
During the turbulent stage, $\gamma_l(m,n)$ is time-dependent (indicated by the $1 \sigma$ spread), and generally much different than $\gamma_s(m,n)$. Significantly,
the turbulent $\gamma_l$ is positive at $n=0$ and low $m$ despite the fact that all linear eigenmodes have $\gamma_s < 0$ for $n=0$. This is a manifestation
of non-normality, for in normal systems, $\gamma_l(m,n) \le \gamma_s(m,n)$. Physically, it is the manifestation of the nonlinear
instability, specifically the linear part of the mechanism in which convective filaments drive density filaments from the equilibrium density gradient.

This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise 
streaks~\cite{trefethen1993,krommes1999}.
Both are transient growth processes. We see this in our simulations by following the evolution of the energy of several $m,n$ modes after turning off the nonlinearities in an already
turbulent simulation. A few representative modes are shown in Fig.~\ref{n0_n1_growth} b). The linear transient growth of the filamentary $n=0$ structures is evident as their modes 
grow transiently before decaying exponentially at the rate of their least stable eigenmode. Such behavior is indicative of non-modal behavior. It must be since 
all $n=0$ linear eigenmodes are stable. Notice also that the $n=1,m=20$ mode decays transiently before growing exponentially with growth rate of the most unstable eigenmode. This transient
decay is also a non-modal result since $\gamma_s(n=1,m=20) > 0$.

Since transient growth is a purely linear phenomenon, it has been a goal of researchers to understand and predict the onset of subcritical turbulence using only linear, non-modal calculations. 
In our system, the turbulence is not subcritical in the traditional sense, but it has a subcritical component in the $n=0$ modes. They have no unstable eigenmodes, but still inject energy into the
turbulence; $\gamma_l(m<50,n=0) > 0$, yet $\gamma_s(m,n=0) < 0$. It is our goal, then, to use linear non-modal calculations to understand this behavior and to move toward predictive capability
of $\gamma_l$.
To accomplish this, we make the anzatz that nonlinearities randomize the turbulent spatial structure at each wavenumber on a time scale of 
one eddy decorrelation time, while the linearities evolve the spatial structures deterministically. From this we can calculate a non-modal $\gamma_l$ spectrum,
which is our prediction of the turbulent $\gamma_l$ spectrum. To illustrate,
we begin by taking Eqs.~\ref{ni_eq}-\ref{te_eq} and Fourier decomposing in the azimuthal and axial directions.  Then, we discritize in the radial direction and
approximate radial derivatives with finite differences.  The resulting system of equations may be written in matrix form:

@@

In order to use non-modal analysis to calculate growth rates and other measures, one must choose a norm and inner product with which to work. While any choice of
inner product is possible, a physically relevant one such as an energy inner product is generally preferred~\cite{camargo1998,schmid2007,camporeale2010}. 
Recall that the inner product of two vectors may be written $\left< \mathbf{x},\mathbf{y} \right> = \mathbf{y}^{\dagger} \mathbf{M} \mathbf{x}$.
We choose $\mathbf{M}$ so that $||\mathbf{v}||^2 = \left< \mathbf{v},\mathbf{v} \right> = E$. Furthermore, it is convenient in
computations to use the $L_2$-norm, $||\mathbf{u}||_2^2 = \sum_i |u_i|^2$. This can be accomplished through the change of variables $\mathbf{u} = \mathbf{M}^{\frac{1}{2}} \mathbf{v}$.
Then the linear portion of Eq.~\ref{mat_eq} becomes

\beq
\label{lin_eq_A}
\diff{\mathbf{u}}{t} = \mathbf{A} \mathbf{u},  \quad \rm{where} \ \mathbf{A} = \mathbf{M}^{\frac{1}{2}} \mathbf{B}^{-1} \mathbf{C} \mathbf{M}^{-\frac{1}{2}}.
\eeq
The solution of Eq.~\ref{lin_eq_A} is $\mathbf{u}(t) = e^{\mathbf{A} t} \mathbf{u}(0)$, which
depends on the initial condition $\mathbf{u}(0)$. For purposes of turbulent growth rate prediction, we are interested in
the behavior of $G(t) = E(t)/E(0) = ||\mathbf{u}(t)||^2/||\mathbf{u}(0)||^2$ and its rate of growth.

@@

It is common practice in normal mode analysis to look for the least stable eigenmode. 
For the non-modal case, it is common to study the properties of $G_{\rm{max}}(t) = ||e^{\mathbf{A} t}||$ because if this is greater than unity at any time, fluctuations may be amplified, 
leading to subcritical turbulence~\cite{trefethen2005,schmid2007}.
However, it can be misleading to study only $G_{\rm{max}}(t)$ when predicting specific properties of turbulence because
$G_{\rm{max}}(t)$ is only the upper envelope of all possible $G(t)$ curves. No one particular initial condition $\mathbf{u}(0)$ evolves along $G_{\rm{max}}(t)$. 
Furthermore, it isn't obvious what kind of spatial structure or structures will come to dominate a turbulent system.
In non-normal systems, unlike in the normal systems, optimal structures don't amplify themselves, rather, they evolve while increasing the total fluctuating energy.

We contend that the key to understanding and prediction turbulent properties through non-modal analysis is to successfully model the effect that the nonlinearities have on the transient linear processes. 
To this effect, we note that the advective nonlinearity in Eq.~\ref{mat_eq} has the form of the state vector divided by a time $\tau_{nl} \sim (v_E k_\perp)^{-1}$. This nonlinear
time scale is generally associated with the eddy turnover or decorrelation time. We therefore present a heuristic model of the nonlinearities 
as a randomizing force that acts on this characteristic nonlinear time scale.
Specifically, we start with an ensemble of random initial conditions, evolve them linearly for a time $\tau_{nl}$, and then take the time and ensemble average growth rate of these curves.
This procedure mimicks what would happen if the turbulence evolved linearly for a time $\tau_{nl}$, randomized, and then repeated. Now, this procedure does not require actual linear
simulations from an ensemble of random initial conditions. Recall that the time evolution of the energy from an initial condition is

@@

\beq
\label{E_t_ensemble_avg}
\left< E(t)/E(0) \right>_{{\rm{ens}}} = \frac{1}{4 N_r} {\rm{tr}} \{ e^{\mathbf{A} t} e^{\mathbf{A}^{\dagger}t} \}.
\eeq
We show the validity of this statistical averaging by plotting $1000$ ensemble curves in Fig.~\ref{m20n0_E_ev} along with their expected average from Eq.~\ref{E_t_ensemble_avg} and their actual average,
which line up perfectly.
One point to note is that in global equation sets like ours, where we discritize and use finite differences in the radial direction rather than a Fourier decomposition, randomizing $\mathbf{u}(0)$
amounts to setting the initial $k_r$ spectrum to a step function that goes to zero at the Nyquist wavenumber. This means that $\left< E(t)/E(0) \right>_{{\rm{ens}}}$ may depend on $N_r$, so we must be careful in
choosing $N_r$ for this analysis. We choose $N_r$ such that the grid spacing equals $\rho_s$, the general Nyquist wavenumber of drift wave simulations~\cite{scott1992}.

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{m20n0_E_ev}}
\caption{An ensemble of initially randomized growth ratio curves (solid gray lines, with the solid black line their average, and the dashed orange line the average as calculated from Eq.~\ref{E_t_ensemble_avg}).
The vertical lines indicate various linear time scales.}
\label{m20n0_E_ev}
\end{figure}
